---
title: "Introduction to R"
output: 
  pdf_document: 
    keep_tex: true
    latex_engine: xelatex
---
``` {r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,
                      fig.align='center',
                      cache=TRUE,
                      out.width = "60%",
                      out.heigth = "60%",
                      warning=FALSE,
                      message=FALSE)
options(width =70)
```
## Exercise 01

Solve the following $Ax = b$ linear system using the 'solve' command:

\begin{center}
A = $\begin{bmatrix}
    12 & -1 & -5 & 0 \\
    -1 & 7 & -2 & -1 \\
    -5 & 2 & 10 & 1 \\
    0 & -1 & 1 & 3
\end{bmatrix}$ b = $\begin{bmatrix}
    1 \\
    2 \\
    3 \\
    4
\end{bmatrix}$

\end{center}

```{r Exercise 1}

a <- matrix(c(12,-1, -5, 0, -1, 7, 2, -1, -5, 2, 10, 1, 0, -1, 1, 3), nrow = 4)
b <- matrix(c(1, 2, 3, 4), nrow = 4)
c <- solve(a, b)
c

```

## Exercise 02
A symmetric matrix $A \in \mathbb{R}^{n \times n}$ (i.e., $A^{t} = A$) is positive definite if, for all nonzero vectors $x \in \mathbb{R}^{n}$, the following condition holds:
\begin{center} $x^{t}Ax > 0$ \end{center}
A criterion used to verify if a symmetric matrix is positive definite is called the "Sylvester's criterion," which states:

\textbf{Sylvester's Criterion:}  $A \in \mathbb{R}^{n \times n}$ is positive definite if and only if $A$ has a positive determinant, and all the submatrices of $A$ listed below have a positive determinant: 
\begin{itemize}
  \item The $1 \times 1$ submatrix formed by the first row and the first column;
  \item The $2 \times 2$ submatrix formed by the first two rows and columns;
  \item \ldots
  \item The $(n - 1) \times (n - 1)$ submatrix formed by the first $(n - 1)$ rows and columns.
\end{itemize}

Write an R function that takes a matrix A as an argument and returns TRUE or FALSE depending on whether it is positive definite or not. Test your function on the matrix A from the previous exercise, which is positive definite.

```{r Exercise 2}
posMatrix <- function(mat) {
  n = nrow(mat)
  m = ncol(mat)
  
  if (n != m) {
    print("The number of rows and columns should be the same!")
  } else {
    if (!all.equal(t(mat), mat)) {
      print("The given matrix is not symmetric")
    } else {
      positive = T
      for (i in 1:n) {
        b = as.matrix(mat[1:i, 1:i])
        determinant = det(b)
        if (determinant < 0) {
          positive = F
          break
        }
      }
      print(positive)
    }
  }   
}

a <- matrix(c(12,-1, -5, 0, -1, 7, 2, -1, -5, 2, 10, 1, 0, -1, 1, 3), nrow = 4)
posMatrix(a)
```


## Exercise 03
Using ggplot2, create plots of the sine and cosine functions for angles ranging from \(-2\pi\) to \(2\pi\). Display both functions in the same figure.

```{r Exercise 3}
library(ggplot2)
library(tidyverse)

df1 <- tibble(column_x = seq(-2*pi, 2*pi, length.out = 1000),
            column_y = sin(column_x),
            fun_type = "f(x) = sin(x)")
            
df2 <- df1 %>%
bind_rows(tibble(column_x = seq(-2*pi, 2*pi, length.out = 1000),
            column_y = cos(column_x),
            fun_type = "f(x) = cos(x)"))


ggplot(df2) +
  aes(x = column_x, y = column_y, color = fun_type) +
  geom_line(size = 1) +
  labs(x = "x", 
  y = "f(x)", 
  title = "Sine and Cosine Functions",
  color = "Function")
```
  

## Exercise 04
We know that solutions to linear systems obtained on a computer, as in Exercise 01, are approximated, since real numbers are represented in an approximate manner on a computer. When the matrix of a linear system is ill-conditioned, the solutions of the linear systems obtained by the computer can have greatly amplified errors. We can quantify the ill-conditioning of an invertible matrix A using the condition number, which is defined as:

\begin{center} $\kappa(A) = \lVert A \rVert \lVert A^{-1} \rVert$ \end{center}

where $\lVert⋅\lVert$ represents the matrix norm, and $A^{-1}$ is the inverse of A. The larger the value of $\kappa(A)$, the more ill-conditioned the matrix is. In other words, solutions of linear systems calculated by computers generally present more numerical errors. You can calculate the condition number in R using the kappa function.

It is possible to improve the conditioning of a symmetric matrix $A \in \mathbb{R}^{n \times n}$ by increasing its diagonal elements. This process is sometimes used in machine learning, for instance, in the regularization process of Ridge Regression. In this exercise, we aim to create a graph illustrating the condition number of A as we increase the elements of its diagonal.

To do this, you can use a symmetric matrix A generated randomly through the following process:
```{r example}
set.seed(1)
X = matrix(runif(100,-1,1),nrow=10)
A = t(X) %*% X
```

In this process, we first generate a matrix $X \in \mathbb{R}^{10 \times 10}$ with random values within the range \([-1, 1]\). Then, we create $A := X^T X$, which results in a symmetric matrix. The set.seed function initializes the seed of the random number generator, ensuring consistent results.

\textbf{Tips:} you can use the \textbf{diag} function to generate an identity matrix. Afterward, calculate the condition number of $A$ with diagonal elements increased by $\lambda$ as follows:
\begin{center} $A' = A + \lambda I$ \end{center}
where $\lambda \geq 0$, and $I \in \mathbb{R}^{d \times d}$ is the identity matrix.

```{r Exercise 4}
library(tidyverse)

cond <- function(mat) {
  I = diag(nrow(mat))
  .data <- tibble(lmbd = numeric(), k = numeric())
  
  for (i in seq(0,1,0.01)) {
    B = A + i*I
    K = kappa(B)
    .data <- add_row(.data, lmbd = i, k = K)
  }
  
  print(.data)
  ggplot(.data) +
    aes(x = lmbd, y = k) +
    geom_line(alpha = 0.3, size = 1.3, color = "red") +
    labs(x = "Lambda", 
    y = "Condition Number", 
    title = "Matrix Conditioning") +
    theme(plot.title = element_text(hjust = 0.5))
}

set.seed(1)
X = matrix(runif(100,-1,1),nrow=10)
A = t(X) %*% X
cond(A)
```