---
title: "Decision Trees"
output: 
  pdf_document: 
    keep_tex: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,
                      fig.align='center',
                      cache=TRUE,
                      out.width = "60%",
                      out.heigth = "60%",
                      warning=FALSE,
                      message=FALSE)
options(width =70)
```

## Exercise 01
Consider the following database composed of four predictor attributes (Outlook, Temperature, Humidity, and Wind) and the Output attribute for classification.
\begin{center}
\begin{tabular}{|c c c c c|} 
 \hline
 Outlook & Temperature & Humidity & Wind & Output \\ [0.5ex] 
 \hline\hline
 Sunny & Hot & High & False & No \\ 
 \hline
 Sunny & Hot & High & True & No \\
 \hline
 Rainy & Cold & Normal & True & No \\
 \hline
 Sunny & Warm & High & False & No \\
 \hline
 Rainy & Warm & High & True & No \\
 \hline
 Overcast & Hot & High & False & Yes \\
 \hline
 Rainy & Warm & High & False & Yes \\
 \hline
 Rainy & Cold & Normal & False & Yes \\
 \hline
 Overcast & Cold & Normal & True & Yes \\
 \hline
 Sunny & Cold & Normal & False & Yes \\
 \hline
 Rainy & Warm & Normal & False & Yes \\
 \hline
 Sunny & Warm & Normal & True & Yes \\
 \hline
 Overcast & Warm & High & True & Yes \\
 \hline
 Overcast & Hot & Normal & False & Yes \\
 \hline
\end{tabular}
\end{center}

Build a decision tree using the CART algorithm and selecting features based on the Gini Impurity.
```{r Exercise 1}
library("rpart.plot")
library("dplyr")   
library("tidymodels")

# Create the dataset
df <- tibble( Outlook = c("Sunny", "Sunny", "Rainy",
                          "Sunny", "Rainy", "Overcast", "Rainy",
                          "Rainy", "Overcast", "Sunny", "Rainy",
                          "Sunny", "Overcast", "Overcast"),
              Temperature = c("Hot", "Hot", "Cold", "Warm", "Warm",
                              "Hot", "Warm", "Cold", "Cold", "Cold",
                              "Warm", "Warm", "Warm", "Hot"),
              Humidity = c("High", "High", "Normal", "High", "High", "High",
                          "High", "Normal", "Normal", "Normal", "Normal",
                          "Normal", "High", "Normal"),
              Windy = c("False", "True", "True", "False",
                        "True", "False", "False", "False", "True",
                        "False", "False", "True", "True", "False"),
              Output = c(rep("No", 5), rep("Yes", 9))) %>%
      mutate_if(is.character, as.factor)
df

# Create the model for the decision tree
tree <- decision_tree( min_n = 5, 
cost_complexity = 0,
tree_depth = 10
) %>%
set_engine("rpart") %>%
set_mode("classification")

# Fit the tree to the data and display the result
t.fit <- tree %>% fit(Output ~.,df)

rpart.plot(t.fit$fit)
```

## Exercise 02
\textbf{Is it possible to improve the performance of a classifier using ensemble methods? What are the advantages of using ensemble methods? What are the main differences between Bagging and Boosting techniques?}

Yes, ensemble methods can enhance the performance of a classifier by combining predictions from multiple models. This results in increased accuracy and greater robustness when compared to using a single model.

The Bagging method is particularly valuable when dealing with models that exhibit high variance, such as decision trees. In this scenario, a series of trees are trained using different subsets of the training data, selected randomly with replacement through a process known as Bootstrapping. The final prediction is obtained by averaging the outputs from these individual trees.

On the other hand, Boosting operates in an incremental manner. Similar to Bagging, it initially takes a sample from the original dataset for training. However, instead of using Bootstrap Sampling, Boosting sequentially trains new models, each one refined based on the errors of the preceding model. This iterative process ultimately yields a more robust predictor. In addition to reducing variance like Bagging, Boosting also contributes to reducing bias.


## Exercise 03
Using the BostonHousing database from the mlbench library, perform a cross-validation process to choose the optimal regularization parameter $\alpha$ (i.e., the cost_complexity parameter). Let min_n=5 and tree_depth=10. Create a plot of the final tree.
```{r Exercise 3}
library("rpart.plot")
library("dplyr")   
library("tidymodels")
library("mlbench")
library("rsample") 
library(tune)

# Load the data
data(BostonHousing)
df <- BostonHousing

# Split the dataset into training and testing sets
set.seed(123)
df_split <- initial_split(df)
df_train <- training(df_split)
df_test  <- testing(df_split)

# Create a model for the decision tree
tune_spec <- 
  decision_tree(
    cost_complexity = tune(),
    tree_depth = 5
  ) %>% 
  set_engine("rpart") %>% 
  set_mode("regression")
tune_spec

# Set up a grid for cross-validation
tree_grid <- grid_regular(cost_complexity(),
                          levels = 5)

set.seed(234)
df_folds <- vfold_cv(df_train)

set.seed(345)
tree_wf <- workflow() %>%
  add_model(tune_spec) %>%
  add_formula(medv ~ .)

tree_res <- 
  tree_wf %>% 
  tune_grid(
    resamples = df_folds,
    grid = tree_grid
    )
tree_res %>% 
  collect_metrics()

# Select the optimal parameters based on the RMSE
best_tree <- tree_res %>%
  select_best("rmse")

# Create the final tree with optimal parameters
final_wf <- 
  tree_wf %>% 
  finalize_workflow(best_tree)

final_fit <- 
  final_wf %>%
  last_fit(df_split) 

final_tree <- extract_workflow(final_fit)

final_tree %>%
  extract_fit_engine() %>%
  rpart.plot(roundint = FALSE)

```

## Exercise 04
Using the BostonHousing database once again, fit the RandomForests and Boosting Trees models. Use the rand_forest and boost_tree commands to define the models, using the randomForest and xgboost engines for each of the algorithms, respectively.

For RandomForest, set a high value for the trees parameter (e.g., 500), set min_n = 5, and set the mtry parameter to √d, where d represents the number of features.

For Boosting Trees, use the same number of trees as in RandomForest for the trees parameter and a learning rate of learn_rate = 0.01. Leave the other parameters at their default values.

Perform cross-validation to estimate the Root Mean Square Error (RMSE) and discuss the results, taking into account the std_err of your results.
```{r Random Forests}
library("rpart.plot")
library("dplyr")   
library("tidymodels")
library("mlbench")
library("rsample") 
library(tune)

## Random Forest

# Load the data
data(BostonHousing)
df <- BostonHousing

# Split the dataset into training and testing sets
set.seed(123)
df_split <- initial_split(df)
df_train <- training(df_split)
df_test  <- testing(df_split)

# Create a model for the Random Forest
tune_spec <- 
  rand_forest(
    trees = 500 ,
    min_n = 5,
    mtry = sqrt(14) 
  ) %>% 
  set_engine("randomForest") %>% 
  set_mode("regression")

# Set up the folds for cross-validation
set.seed(234)
df_folds <- vfold_cv(df_train)

set.seed(345)
tree_wf <- workflow() %>%
  add_model(tune_spec) %>%
  add_formula(medv ~ .)

# Apply cross-validation
tree_res <- 
  tree_wf %>% 
  tune_grid(resamples = df_folds)

# Collect metrics
metrics <- tree_res %>% 
  collect_metrics()
print(metrics)
```

```{r boost tree}
library("rpart.plot")
library("dplyr")   
library("tidymodels")
library("mlbench")
library("rsample") 
library(tune)

## Boosting Tree

# Load the data
data(BostonHousing)
df <- BostonHousing

# Split the dataset into training and testing sets
set.seed(123)
df_split <- initial_split(df)
df_train <- training(df_split)
df_test  <- testing(df_split)

# Create a model for the boost tree
tune_spec <- 
  boost_tree(
    trees = 500 ,
    learn_rate = 0.01
  ) %>% 
  set_engine("xgboost") %>% 
  set_mode("regression")

# Set up the folds for cross-validation
set.seed(234)
df_folds <- vfold_cv(df_train)

set.seed(345)
tree_wf <- workflow() %>%
  add_model(tune_spec) %>%
  add_formula(medv ~ .)

# Apply cross-validation
tree_res <- 
  tree_wf %>% 
  tune_grid(resamples = df_folds)

# Collect metrics
metrics <- tree_res %>% 
  collect_metrics()
print(metrics)
```

## Exercise 05
We have seen that Random Forests, like Bagging-based methods, can be used to identify the most important features. Explain how these features are identified and calculate the most important variables within the BostonHousing dataset. Identify the three most important features according to this criterion.
```{r Exercise 5}
library("rpart.plot")
library("dplyr")   
library("tidymodels")
library("mlbench")
library("rsample") 
library("vip")

# Load the data
data(BostonHousing)
df <- BostonHousing

# Split the dataset into training and testing sets
set.seed(123)
df_split <- initial_split(df)
df_train <- training(df_split)
df_test  <- testing(df_split)

# Create a model for the random forest
tune_spec <- 
  rand_forest(
    trees = 500 ,
    min_n = 5,
    mtry = sqrt(14),
  ) %>% 
  set_engine("randomForest") %>% 
  set_mode("regression")
tune_spec

t.fit <- tune_spec %>% fit(medv ~.,df)

# Check the most relevant features
vip(t.fit)
```
The most relevant variables can be obtained by ranking the predictors with the highest average reduction of RSS (Residual Sum of Squares) or Gini index across the built trees. In this case, the top 3 predictor variables are those with the highest values along the X-axis of the above graph, namely rm, lstat, and nox.